Definition of Archipelagos

### 2025-01-31 application_examples/abc_inferences/archipelago-definition.Rmd
Updating state of branch develop to checkout to cftp

# Initialize results list
#results_list <- vector("list", nrow(region_gff))
## TODO: Update comment
cgiVsDistance <- matrix(data = NA, nrow = nrow(genome_gff), ncol = 20000)

@@

>>>>

# Initialize result dataframe
cgiVsDistance <- matrix(data = NA, nrow = nrow(genome_gff), ncol = 20000)

@@

## Step 3: Island assignation to closer gene

Neglecting those islands that intersect gene bodies, we get the overlaps of CGI with upstream regions:

```{r}
library(GenomicRanges)

### Step 1: Filter CGI not within gene bodies ###

# Create GRanges for two sets with different chromosomes
gr_cgi <- GRanges(seqnames = cgif$chr,
                  ranges = IRanges(start = cgif$start,
                                   end = cgif$end))

gr_genes <- GRanges(seqnames = genome_gff$seqnames,
                    ranges = IRanges(start = genome_gff$start,
                                     end = genome_gff$end))

# Subset common chromosomes (seqnames)
common_seqnames <- intersect(seqlevels(gr_cgi), seqlevels(gr_genes))
gr_cgi <- keepSeqlevels(gr_cgi, common_seqnames, pruning.mode = "coarse")
gr_genes <- keepSeqlevels(gr_genes, common_seqnames, pruning.mode = "coarse")

# Find overlaps
overlap_result <- overlapsAny(gr_cgi, gr_genes)

# Subset non-overlapping CG islands
non_overlapping_cgi <- gr_cgi[!overlap_result]

### Step 2: Find overlaps of CGI with upstream regions ###

# Filter out rows with NA in upstream_start or upstream_end 
genome_gff_clean <- genome_gff[!is.na(genome_gff$upstream_start) & !is.na(genome_gff$upstream_end), ]

# Rename upstream_start and upstream_end from the "-" strand as in GRanges objects
# each range must have an end that is greater or equal to its start minus one
inverted_start <- genome_gff_clean$upstream_end[genome_gff_clean$strand == "-"]
inverted_end <- genome_gff_clean$upstream_start[genome_gff_clean$strand == "-"]
genome_gff_clean$upstream_start[genome_gff_clean$strand == "-"] <- inverted_start
genome_gff_clean$upstream_end[genome_gff_clean$strand == "-"] <- inverted_end

# Create a GRanges object for upstream regions
upstream_regions <- GRanges(
  seqnames = genome_gff_clean$seqnames,
  ranges = IRanges(start = genome_gff_clean$upstream_start, end = genome_gff_clean$upstream_end),
  gene_id = genome_gff_clean$ID,
  strand = genome_gff_clean$strand
)

# Subset common chromosomes (seqnames)
common_seqnames <- intersect(seqlevels(non_overlapping_cgi), seqlevels(upstream_regions))
non_overlapping_cgi <- keepSeqlevels(non_overlapping_cgi, common_seqnames, pruning.mode = "coarse")
upstream_regions <- keepSeqlevels(upstream_regions, common_seqnames, pruning.mode = "coarse")

# Find overlaps between non-overlapping CG islands and upstream regions
upstream_overlap <- findOverlaps(non_overlapping_cgi, upstream_regions)

### Step 3: Extract details of overlaps ###

# Query (CGI) and subject (upstream regions) indices
cgi_indices <- queryHits(upstream_overlap)
upstream_indices <- subjectHits(upstream_overlap)

# Calculate the number of overlaps for each CGI
cgi_overlap_counts <- table(cgi_indices) # Creates a named vector with CGI indices as names and counts as values

# Create a vector that maps the overlap count to each CGI in overlap_info
overlap_number <- cgi_overlap_counts[as.character(cgi_indices)] # Map counts back to the corresponding CGI index

# Map upstream regions back to the original genome_gff_clean to get gene start and end
gene_start <- genome_gff_clean$start[match(mcols(upstream_regions)$gene_id[upstream_indices], genome_gff_clean$ID)]
gene_end <- genome_gff_clean$end[match(mcols(upstream_regions)$gene_id[upstream_indices], genome_gff_clean$ID)]

# Combine relevant information
overlap_info <- data.frame(
  CGI_Chr = seqnames(non_overlapping_cgi)[cgi_indices],
  CGI_Start = start(non_overlapping_cgi)[cgi_indices],
  CGI_End = end(non_overlapping_cgi)[cgi_indices],
  Gene_ID = mcols(upstream_regions)$gene_id[upstream_indices],
  Gene_Start = gene_start,
  Gene_End = gene_end,
  Upstream_Start = start(upstream_regions)[upstream_indices],
  Upstream_End = end(upstream_regions)[upstream_indices],
  Upstream_Length = width(upstream_regions)[upstream_indices],
  Strand = as.character(strand(upstream_regions)[upstream_indices]),
  Overlap_Number = as.integer(overlap_number)
)

# Calculate the distance to the gene based on the strand
distance_to_gene <- ifelse(
  overlap_info$Strand == "+",
  overlap_info$Gene_Start - overlap_info$CGI_End, # Positive strand: distance from CGI_End to Gene_Start
  overlap_info$CGI_Start - overlap_info$Gene_End # Negative strand: distance from CGI_Start to Gene_End
)

# Add the distance_to_gene column to overlap_info
overlap_info$Distance_to_Gene <- distance_to_gene

# Inspect the result
head(overlap_info)

Number of upstream regions each island intersects with:

```{r}
table(overlap_info$Overlap_Number)
```

Summary of distances from the islands to the associated gene:

```{r}
summary(overlap_info$Distance_to_Gene)
```

```{r}
# Count the number of CGIs assigned to each gene
gene_cgi_counts <- table(overlap_info$Gene_ID)

# Create a frequency table of these counts
count_distribution <- table(gene_cgi_counts)

# Plot
barplot(
  height = as.numeric(count_distribution), # y-axis: number of genes
  names.arg = names(count_distribution),   # x-axis: CGI counts per gene
  xlab = "Number of CGIs assigned to a gene", 
  ylab = "Number of genes",
  main = "Distribution of CGI Assignments to Genes",
  col = "skyblue",
  border = "black"
)
```

Now from those islands that were associated to 2 upstream regions (positive and negative strand), we keep only the entry corresponding to the closer gene:

```{r}
# Step 1: Split overlap_info by unique CGI coordinates
cgi_groups <- split(overlap_info, paste(overlap_info$CGI_Chr, overlap_info$CGI_Start, overlap_info$CGI_End))

# Step 2: For each group, keep the row with the smallest Distance_to_Gene
filtered_overlap_info <- do.call(rbind, lapply(cgi_groups, function(group) {
  group[which.min(group$Distance_to_Gene), ] # Select the row with the minimum Distance_to_Gene
}))

# Step 3: Reset row names
rownames(filtered_overlap_info) <- NULL

# Step 4: Inspect the filtered result
head(filtered_overlap_info)

```{r}
# Check that the number of entries filtered out corresponds to half of the original number of entries corresponding to islands assigned to 2 genes
nrow(overlap_info) - nrow(filtered_overlap_info) == table(overlap_info$Overlap_Number)[2]/2
```

```{r}
# Count the number of CGIs assigned to each gene
gene_cgi_counts <- table(filtered_overlap_info$Gene_ID)

# Create a frequency table of these counts
count_distribution <- table(gene_cgi_counts)

# Plot
barplot(
  height = as.numeric(count_distribution), # y-axis: number of genes
  names.arg = names(count_distribution),   # x-axis: CGI counts per gene
  xlab = "Number of CGIs assigned to a gene", 
  ylab = "Number of genes",
  main = "Distribution of CGI Assignments to Genes",
  col = "skyblue",
  border = "black"
)
```

**Maybe I should check the order of the filtering, as all those genes overlapping with each other were removed from genome_gff before, so maybe some of the islands that are kept here because they dont overlap gene bodies actually would overlap with those genes that were removed**

*Not now at least, but keep it in mind when looking at the results*
